\documentclass[10pt]{article}

% Venue-neutral source. Replace the class and author block when a target venue
% is selected; keep the evidence hierarchy and claim boundaries intact.
\usepackage[margin=1in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{url}

\newcommand{\lcp}{LCP-style soft penalty}
\newcommand{\scppo}{SC-PPO 3.8 PID}
\newcommand{\heur}{Revised heuristic}
\newcommand{\jacc}{joint-acceleration}

\title{Policy Local Sensitivity Regularization for Smooth Humanoid Locomotion}
\author{Anonymous Authors}
\date{}

\begin{document}
\maketitle

\begin{abstract}
Smooth humanoid locomotion is often pursued by adding action-rate or
torque-rate penalties to the reward. These penalties are practical, but they
regularize realized behavior only indirectly and require task-specific scalar
weight tuning. We study policy local sensitivity as a mechanism-level object
for smooth control: instead of only penalizing the motion that a policy has
already produced, we regulate how strongly the policy map can amplify
observation perturbations into action changes. The paper compares three
evidence lines under a shared Humanoid-Gym rough-terrain protocol: SC-PPO, a
hard policy-Jacobian constraint with a PID-Lagrangian multiplier; an LCP-style
soft Jacobian/Lipschitz penalty; and a revised action-rate heuristic. On a
five-seed Isaac Gym audit, the LCP-style penalty is the strongest current
local-sensitivity row, with zero fall rate, lower velocity error, lower action
jitter, higher return, and much lower measured policy sensitivity than SC-PPO.
SC-PPO remains useful as the hard-constraint mechanism that exposes
enforcement and seed-sensitivity trade-offs. In matched five-seed MuJoCo replay,
the LCP-style row has the lowest action jitter and is much stronger than
SC-PPO, while the revised heuristic remains better on aggregate joint
acceleration and return. This mixed MuJoCo result is not a contradiction:
local sensitivity primarily regularizes the policy-output stream, while joint
acceleration and return are downstream closed-loop outcomes involving tracking,
contact timing, PD response, and simulator dynamics. The resulting claim is
mechanism-level rather than a single-method dominance claim:
policy-local-sensitivity regularization is a useful smooth-control lens, but
enforcement details matter and no single row dominates every metric.
\end{abstract}

\section{Introduction}

Reinforcement-learning locomotion policies can satisfy task reward while
producing dynamically rough behavior. In humanoid locomotion this appears as
high-frequency action variation, large joint accelerations, and policies whose
smoothness depends on simulator details. A policy can look acceptable in its
training engine yet respond sharply to small observation or contact-solver
differences when replayed in another physics engine.

The standard engineering response is reward shaping. Action-rate, torque-rate,
joint-acceleration, and similar penalties are easy to add and can work well in
practice. The cost is that the smoothness trade-off becomes a scalar tuning
problem. The policy is not directly constrained to be locally stable as a map
from observations to actions; it is only penalized after it has already emitted
actions.

This paper asks whether policy local sensitivity is a useful mechanism for
smooth humanoid control. The central object is the local derivative of policy
output with respect to policy observations. The mechanism hypothesis is simple:
if a policy strongly amplifies small observation differences into action
differences, then simulator-specific contact and integration differences can
become action jitter and dynamic roughness. Regulating that amplification should
therefore help smoothness, but different enforcement mechanisms may have
different robustness.

The evidence does not support a simple ``SC-PPO wins'' story. Instead, the
strongest current result is a mechanism comparison. A soft LCP-style
Jacobian/Lipschitz penalty is more robust than the current SC-PPO hard-constraint
implementation. The revised heuristic remains a strong anchor, especially in
matched MuJoCo joint acceleration and return. SC-PPO remains scientifically
useful because it exposes how a hard policy-Jacobian constraint with
PID-Lagrangian enforcement behaves under PPO, seed variation, checkpoint
selection, and cross-engine replay. OmniSafe PPO-Lag is retained as a
framework-interface diagnostic, not as a promoted external constrained-RL
baseline.

The contributions are:
\begin{enumerate}
  \item We formulate and audit policy local sensitivity as a smooth-control
  mechanism for humanoid locomotion.
  \item We compare hard policy-sensitivity enforcement, soft
  LCP-style Jacobian/Lipschitz regularization, and a revised action-rate
  heuristic under the same five-seed Isaac Gym protocol.
  \item We add matched five-seed MuJoCo replay and show a metric split:
  LCP-style regularization is cleanest on action jitter, while the heuristic
  remains better on aggregate joint acceleration and return.
  \item We report robustness audits covering paired bootstrap uncertainty,
  selected-vs-final checkpoint dependence, local LCP coefficient sensitivity,
  and a controlled policy perturbation audit.
  \item We document framework-boundary diagnostics for OmniSafe PPO-Lag and
  local CPO-style updates without promoting either line to a main baseline.
\end{enumerate}

\section{Related Work}

Constrained RL separates reward maximization from costs or limits. CPO
formulates policy optimization under explicit constraints~\citep{Achiam2017CPO},
while Safety Gym and PPO-Lagrangian practice provide common infrastructure for
cost-aware policy learning~\citep{Ray2019SafetyGym}. PID-Lagrangian methods
address multiplier overshoot and responsiveness in constrained deep
RL~\citep{Stooke2020PID}. OmniSafe packages many safe-RL algorithms behind a
common infrastructure~\citep{Ji2024OmniSafe}. Recent humanoid work also uses
constrained optimization to regulate physical costs such as
energy~\citep{Huang2026ECO}. This paper belongs to that cost/reward separation
lineage, but the cost placement is different: the SC-PPO cost is actor-internal
and depends on the derivative of the current policy with respect to observations
during PPO updates.

Smoothness in learned locomotion is commonly encouraged through reward terms on
action rate, torque rate, or acceleration. These are practical and strong
baselines; the revised heuristic in this paper is therefore treated as a
reward-shaping anchor, not as a strawman. A different family regularizes the
policy map itself. Spectral normalization constrains network Lipschitz
constants~\citep{Miyato2018SN}, and LayerNorm changes hidden activation
scaling~\citep{Ba2016LayerNorm}. LCP directly targets smooth humanoid
locomotion through a Lipschitz/Jacobian-style policy
regularizer~\citep{Chen2024LCP}, with adjacent work exploring spectral-normalization
variants for Lipschitz-constrained policies~\citep{Shin2025SNLCP}. Our LCP row
should be read as a local same-task LCP-style adaptation under Humanoid-Gym
rough terrain, not as official LCP checkpoint or code parity.

Cross-engine replay is intermediate validation, not hardware validation. Isaac
Gym provides high-throughput GPU simulation for robot
learning~\citep{Makoviychuk2021IsaacGym}, while MuJoCo provides a different physics
engine and contact implementation~\citep{Todorov2012MuJoCo}. Humanoid-Gym
supplies the training scaffold and zero-shot transfer
context~\citep{Gu2024HumanoidGym}. We use Isaac Gym for training/evaluation and
aligned MuJoCo replay as a stress test for whether dynamic smoothness survives
simulator changes. We do not claim real-robot transfer.

\section{Methods}

\subsection{Shared PPO Task}

All primary rows use the same Humanoid-Gym rough-terrain locomotion task and
PPO backbone~\citep{Schulman2017PPO}. The robot is an H1-class humanoid with
12 DoF. The full-paper audit uses seeds 11, 17, 23, 29, and 31, a
512-environment training budget, and a checkpoint sweep over the trained policy.
Evaluation reports fall rate, velocity-tracking error, joint-acceleration
$L_2$, action-jitter $L_2$, episode return, and policy-sensitivity side reads
where available.

The selected-checkpoint protocol is explicit: checkpoints are filtered by task
validity and then selected by the smoothness-first rule used throughout the
repository. Final-checkpoint behavior is reported separately in the checkpoint
robustness audit.

\subsection{SC-PPO Hard Policy-Sensitivity Constraint}

SC-PPO introduces a constraint cost based on policy local sensitivity. For a
policy mean $\mu_\theta(o)$, the local cost is the Frobenius norm of the
observation-to-action Jacobian:
\begin{equation}
  C(o) = \left\| \frac{\partial \mu_\theta(o)}{\partial o} \right\|_F .
\end{equation}
The batch cost for multiplier updates is a tail-sensitive statistic:
\begin{equation}
  J_C(\theta) = Q_{0.90}\left(\{C(o_i)\}\right).
\end{equation}
SC-PPO optimizes a Lagrangian objective
\begin{equation}
  \mathcal{L}(\theta,\lambda) =
  J_R(\theta) - \lambda \left(J_C(\theta) - d\right),
\end{equation}
with threshold $d=3.8$, $\lambda_{\mathrm{init}}=0.5$, quantile aggregation,
and \texttt{lower\_bound\_clamp} PID integral handling. Heuristic smoothness
rewards are disabled in the SC-PPO row, so smoothness pressure comes from the
Jacobian constraint rather than reward shaping.

\subsection{LCP-Style Soft Jacobian/Lipschitz Penalty}

The LCP-style row uses a fixed soft penalty on the policy-gradient quantity
$\|\nabla_o \log \pi_\theta(a \mid o)\|^2$, following the mechanism family of
Lipschitz-constrained policies~\citep{Chen2024LCP}. It uses
\texttt{lcp\_weight = 0.002}, does not use a Lagrange multiplier, and disables
heuristic smoothness rewards. It shares the same task, checkpoint sweep, metric
schema, and MuJoCo replay bridge as SC-PPO and the heuristic.

This is the closest literature-aligned policy-sensitivity baseline in the
current repository, but it is not an official LCP reproduction. The public
LCP/MimicKit stack uses a different task, robot, checkpoint setup, and
evaluation path; the paper therefore labels this row LCP-style rather than
claiming official parity.

\subsection{Revised Heuristic Reward-Shaping Anchor}

The heuristic row is PPO with a tuned action-rate penalty under the formal
protocol revision. It represents the practical reward-shaping approach that the
policy-sensitivity rows must beat or explain. Because it remains highly
competitive in MuJoCo, especially on joint acceleration and return, it is a
strong anchor rather than a weak baseline.

\subsection{OmniSafe PPO-Lag and CPO Diagnostics}

OmniSafe PPO-Lag was evaluated as a bounded migration diagnostic. The adapter,
cost bridge, policy evaluation bridge, and update-hook smoke paths were
implemented, but the three-seed end-to-end diagnostic collapsed. This is not
reported as a promoted external constrained-RL baseline. The result shows that
a standard environment-side PPO-Lag interface is not a faithful drop-in route
for this actor-internal Jacobian cost without algorithm-level hooks.

A local CPO-style path was also tested only as future-work evidence. Tensor and
one-update checks passed, and a bounded training-loop diagnostic produced
finite accepted updates, but all evaluated checkpoints collapsed. We therefore
do not include CPO in the main baseline table.

\section{Experimental Protocol}

The primary Isaac comparison uses five seeds, checkpoint sweeps, and the same
metric schema for LCP-style, SC-PPO, and heuristic rows. MuJoCo replay uses the
selected Isaac checkpoint for each seed under the aligned \texttt{isaac\_mainline}
protocol with 20 episodes of 20 seconds and \texttt{joint\_reset\_noise = 0.1}.

Metrics are fall rate, velocity-tracking error, joint-acceleration $L_2$,
action-jitter $L_2$, episode return, and policy sensitivity. Fall, velocity
error, joint acceleration, action jitter, and sensitivity are lower-is-better;
episode return is higher-is-better. Policy sensitivity and violation rate are
side reads for policy-local behavior rather than task reward.

The evidence tiers are: primary full-paper rows (LCP-style soft penalty,
SC-PPO 3.8 PID-Lagrangian, and revised heuristic), robustness audits (paired
bootstrap uncertainty, selected-vs-final checkpoint audit, and LCP coefficient
sensitivity), and diagnostics (policy perturbation audit, OmniSafe PPO-Lag
migration, local CPO-style checks, historical workshop-era SC-PPO evidence, and
other closed alternative mechanisms).

\section{Results}

\subsection{Five-Seed Isaac Audit}

Table~\ref{tab:isaac} shows the primary five-seed Isaac Gym
selected-checkpoint comparison. The LCP-style soft penalty is the strongest
current local-sensitivity row. It has zero fall rate, lower velocity error,
lower action jitter, higher return, and much lower measured policy sensitivity
than SC-PPO. The heuristic remains competitive and is slightly better on Isaac
joint acceleration.

\begin{table}[t]
\centering
\caption{Five-seed Isaac Gym selected-checkpoint comparison. Values are
mean $\pm$ standard deviation over seeds 11/17/23/29/31.}
\label{tab:isaac}
\small
\resizebox{\linewidth}{!}{%
\begin{tabular}{lccccccc}
\toprule
Method & Checkpoints & Fall & Vel. err & Jnt acc & Jitter & Return & Sens. \\
\midrule
\lcp & 300/400/400/400/400 & $0.000\pm0.000$ & $0.490\pm0.032$ & $117.331\pm21.171$ & $0.212\pm0.033$ & $118.420\pm6.595$ & $1.890\pm0.064$ \\
\scppo & 300/300/400/400/400 & $0.170\pm0.194$ & $0.606\pm0.087$ & $142.955\pm52.454$ & $0.277\pm0.100$ & $99.349\pm13.959$ & $3.630\pm0.067$ \\
\heur & 350/300/350/400/400 & $0.150\pm0.063$ & $0.705\pm0.103$ & $115.317\pm8.770$ & $0.260\pm0.018$ & $105.326\pm10.335$ & $7.331\pm0.237$ \\
\bottomrule
\end{tabular}}
\end{table}

The paired bootstrap audit supports the conservative statement that LCP is
stronger than SC-PPO on Isaac fall, velocity error, return, and measured
sensitivity. The joint-acceleration advantage over SC-PPO is directionally
favorable but not clean under five seeds because the confidence interval
overlaps zero. LCP also beats the heuristic on fall, velocity error, jitter,
return, and sensitivity, while heuristic joint acceleration remains
metric-competitive.

\subsection{Matched MuJoCo Replay}

Table~\ref{tab:mujoco} reports matched five-seed MuJoCo selected replay. The
LCP-style row has the lowest action jitter and is much stronger than SC-PPO.
The revised heuristic remains better on aggregate joint acceleration and
return.

\begin{table}[t]
\centering
\caption{Matched five-seed MuJoCo selected-checkpoint replay. Values are
mean $\pm$ standard deviation over the same selected checkpoints as
Table~\ref{tab:isaac}.}
\label{tab:mujoco}
\small
\resizebox{\linewidth}{!}{%
\begin{tabular}{lccccc}
\toprule
Method & Fall & Vel. err & Jnt acc & Jitter & Return \\
\midrule
\lcp & $0.000\pm0.000$ & $0.406\pm0.060$ & $117.425\pm11.234$ & $0.195\pm0.029$ & $-599.108\pm236.114$ \\
\scppo & $0.010\pm0.020$ & $0.471\pm0.079$ & $159.718\pm71.345$ & $0.322\pm0.167$ & $-627.238\pm222.350$ \\
\heur & $0.000\pm0.000$ & $0.406\pm0.043$ & $111.615\pm14.597$ & $0.226\pm0.037$ & $-456.370\pm195.030$ \\
\bottomrule
\end{tabular}}
\end{table}

This result should be read as a control-path metric split. Policy-local
sensitivity regularization directly targets the policy-output stream, which is
consistent with LCP winning action jitter on aggregate and on three of five
seeds. Joint acceleration and return are downstream closed-loop outcomes that
also depend on velocity tracking, contact timing, PD target dynamics, and
simulator-specific response. The leave-one-seed audit preserves the split:
LCP remains the best aggregate action-jitter row in every split, while the
heuristic remains best on aggregate joint acceleration and return in every
split.

\subsection{Policy Perturbation Mechanism Audit}

A controlled policy perturbation audit adds a more direct mechanism check.
On a shared observation bank from selected-checkpoint rollouts, deterministic
action responses to small observation perturbations follow the expected
ordering: LCP-style soft penalty has the lowest amplification, SC-PPO is
intermediate, and the revised heuristic is largest.

\begin{table}[t]
\centering
\caption{Policy perturbation audit on a shared observation bank. Amplification
is $\|\Delta a\|/\|\Delta o\|$ at perturbation radius $0.01$ using deterministic
action means.}
\label{tab:perturb}
\small
\resizebox{\linewidth}{!}{%
\begin{tabular}{lccccccc}
\toprule
Method & Seeds & Amp. mean & Amp. p90 & Bank sens. & Selected sens. & Jitter & Jnt acc \\
\midrule
\lcp & 5 & 0.082 & 0.133 & 2.308 & 1.890 & 0.212 & 117.331 \\
\scppo & 5 & 0.153 & 0.241 & 4.307 & 3.630 & 0.277 & 142.955 \\
\heur & 5 & 0.289 & 0.468 & 8.341 & 7.331 & 0.260 & 115.317 \\
\bottomrule
\end{tabular}}
\end{table}

Across the 15 method-seed rows, amplification mean correlates with selected
policy sensitivity at $0.999$, but only weakly with selected action jitter
($0.228$) and selected joint acceleration ($-0.093$). This supports the local
policy-output mechanism without turning downstream joint acceleration or return
into direct sensitivity metrics.

\subsection{Robustness Audits}

The statistical robustness audit is descriptive rather than a large-sample
significance test. With five seeds, it supports stable directions and
uncertainty intervals, not binary claims. The strongest robust statement is:
LCP is clearly stronger than the current SC-PPO hard-constraint row on Isaac
fall, velocity error, return, sensitivity, and MuJoCo action jitter.

The selected-vs-final checkpoint audit shows that LCP is close to final-only
behavior. Only seed 11 selects checkpoint 300 instead of final checkpoint 400,
and selected-to-final aggregate deltas are small. SC-PPO is more
dynamic-selection-sensitive: final checkpoints improve velocity and return but
worsen joint acceleration and jitter. The heuristic is task-selection-sensitive:
final checkpoints improve velocity and return slightly but increase fall rate
and dynamic roughness.

The LCP coefficient diagnostic tests weights 0.001, 0.002, and 0.004 on seeds
23/29/31. The middle value, 0.002, is best on the selected aggregate for fall,
velocity error, joint acceleration, action jitter, and return. This does not
prove global hyperparameter optimality, but it shows that the main LCP row is
not an isolated single-point accident in the immediate neighborhood.

\section{Discussion}

\subsection{What the Mechanism Comparison Shows}

The evidence supports policy local sensitivity as a useful smooth-control lens,
but it does not support hard-constraint SC-PPO as the strongest implementation.
The LCP-style soft penalty is currently stronger under the same task and metric
schema. This suggests that the regulated object is useful while the enforcement
mechanism matters.

Soft regularization may be easier to optimize in PPO because the penalty is
present throughout actor updates with a fixed coefficient. SC-PPO exposes the
cost of hard enforcement: the constraint is interpretable, but the PID
multiplier and checkpoint behavior are seed-sensitive. The contribution of
SC-PPO is therefore not final performance dominance; it is the mechanistic
contrast between hard constraint enforcement and soft policy-map regularization.

\subsection{Why the Heuristic Still Matters}

The revised heuristic should remain central in the paper. It is not a weak
baseline. In matched MuJoCo replay it is better than LCP on aggregate joint
acceleration and return, even though LCP is better on action jitter. This makes
the paper stronger, not weaker: the mechanism story must explain a real
trade-off rather than dismiss reward shaping.

The current reading is that action jitter is closest to the policy-output
stream, while joint acceleration and return depend on the full closed loop. A
reward-shaped policy can sometimes produce better downstream dynamics even if
its policy map is more sensitive. Conversely, a low-sensitivity policy can
produce cleaner action commands without winning every downstream metric.

\subsection{Sim-to-Sim Scope}

MuJoCo replay is valuable because it changes physics-engine details without
changing the trained policy checkpoint. It is not hardware validation. The
sim-to-sim result supports the idea that policy-output smoothness and dynamic
smoothness can be stress-tested across engines, but it does not establish
sim-to-real robustness. Domain randomization, system identification, actuator
modeling, and hardware tests remain outside the current evidence.

\section{Limitations}

The LCP-style row is a local same-task adaptation, not official LCP
code/checkpoint parity. The paper should not say official LCP is reproduced or
beaten.

The primary evidence uses one robot morphology, one rough-terrain task, five
training seeds, and selected-checkpoint evaluation. The selected-vs-final audit
reduces this concern for LCP but does not eliminate checkpoint dependence for
all methods.

MuJoCo replay is sim-to-sim evidence only. There is no real-robot validation,
calibrated actuator model, system identification, or multi-terrain/multi-robot
study. The actuator low-pass proxy stress test is diagnostic and should not be
written as hardware evidence.

The mechanism explanation remains local and partly correlational. The policy
perturbation audit gives controlled local policy-output evidence, and the
matched MuJoCo decomposition supports a control-path metric split, but neither
proves a closed-loop causal intervention from sensitivity to joint acceleration
or return.

The statistical audit is a five-seed uncertainty audit, not a large-sample
hypothesis test. Confidence intervals and bootstrap rank frequencies should be
reported as descriptive robustness evidence.

External constrained-RL coverage remains incomplete. OmniSafe PPO-Lag was
tested as a bounded interface diagnostic and collapsed, but this does not show
that OmniSafe, PPO-Lag, CPO, or external constrained RL broadly fail. A local
CPO-style path was also tested through autograd/HVP, one-update, and bounded
training-loop diagnostics. The tensor and one-update checks passed, and the
training loop produced finite accepted updates, but every evaluated checkpoint
collapsed. CPO is therefore future work rather than a current baseline.

\section{Conclusion}

This paper should not claim that SC-PPO is the strongest current method, and it
should not claim that LCP dominates every metric. The defensible result is more
specific: policy-local-sensitivity regularization is a useful mechanism for
smooth humanoid control, and the enforcement mechanism matters. A fixed soft
LCP-style Jacobian/Lipschitz penalty is the strongest current local-sensitivity
row under the same Humanoid-Gym protocol. SC-PPO remains valuable because it
shows how a hard policy-Jacobian constraint behaves under PID-Lagrangian
enforcement, seed variation, checkpoint selection, and framework migration. The
revised heuristic remains a strong reward-shaping anchor, especially in MuJoCo
joint acceleration and return. OmniSafe PPO-Lag clarifies a framework boundary:
actor-internal Jacobian costs require algorithm-level access and should not be
reduced to environment-side proxy costs. Together, these results support a
mechanism-comparison paper about local sensitivity, not a broad dominance or
hardware-transfer claim.

\bibliographystyle{plainnat}
\bibliography{references}

\appendix

\section{Diagnostic Tables}

\begin{table}[htbp]
\centering
\caption{Local LCP coefficient diagnostic on seeds 23/29/31.}
\label{tab:lcp-weight}
\small
\resizebox{\linewidth}{!}{%
\begin{tabular}{lcccccc}
\toprule
Weight & Fall & Vel. err & Jnt acc & Jitter & Return & Sens. \\
\midrule
0.001 & 0.017 & 0.569 & 147.380 & 0.267 & 107.706 & 2.241 \\
0.002 & 0.000 & 0.477 & 123.903 & 0.226 & 119.697 & 1.861 \\
0.004 & 0.000 & 0.539 & 128.041 & 0.237 & 117.982 & 1.712 \\
\bottomrule
\end{tabular}}
\end{table}

\begin{table}[htbp]
\centering
\caption{OmniSafe PPO-Lag diagnostic. Collapsed policies are not smooth
locomotion evidence; low smoothness metrics in the selected row come from
falling policies.}
\label{tab:omnisafe}
\small
\resizebox{\linewidth}{!}{%
\begin{tabular}{lccccc}
\toprule
Slice & Fall & Vel. err & Jnt acc & Jitter & Return \\
\midrule
Selected diagnostic & 1.000 & 1.468 & 79.309 & 0.008 & 4.386 \\
Final checkpoint 400 & 1.000 & 1.109 & 321.328 & 0.741 & 2.714 \\
\bottomrule
\end{tabular}}
\end{table}

\section{Evidence Map}

The primary paper-facing evidence consists of the five-seed Isaac comparison,
the matched five-seed MuJoCo selected replay, the policy perturbation audit,
the paired bootstrap uncertainty audit, selected-vs-final checkpoint
robustness, and local LCP coefficient sensitivity. OmniSafe PPO-Lag and local
CPO-style results are retained as diagnostic or future-work evidence only.

The older workshop-era SC-PPO-centered draft should be treated as historical
context for the project trajectory. It used a narrower three-seed
SC-PPO-vs-heuristic frame and did not include the current LCP-style full-paper
baseline, matched five-seed MuJoCo replay, statistical audit, checkpoint
robustness audit, policy perturbation audit, or related-work claim-boundary
map.

\end{document}
